Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-23 paper.qmd
Update paper.qmd
formatting

Other packages provide similar functionality with different trade-offs.
[groundhog]{.pkg} by @simonsohn2023 makes it easy to install packages as they
existed on CRAN at a given date:
```{=tex}
\begin{CodeInput}
R> groundhog.library("
+   library('purrr')
+   library('ggplot2')",
+   "2017-10-04",
+   tolerate.R.version = "4.2.2")
\end{CodeInput}
```

@@

5. Incomplete Reproducibility: Reproducibility should be understood as a spectrum
  rather than a binary property, a notion emphasised in "Reproducibility in Software
  Engineering" by @dellaieraMasterThesis2024. Docker illustrates this spectrum well:
  it provides strong run-time reproducibility—that is, consistency in space
  (where the same image behaves identically across different systems) but weak
  build-time reproducibility, which concerns consistency over time. As defined in
  the thesis, reproducibility at build time means being able to produce an identical
  build artefact regardless of when or where it is built. However, Docker inherently
  fails to guarantee this property. Its reliance on mutable base images and non-deterministic
  instructions such as `apt-get update` leads to temporal drift: rebuilding the same
  `Dockerfile` at different points in time can produce distinct outputs. In the
  experimental evaluation conducted in the thesis, even with pinned base images,
  checksums of the resulting containers varied across builds and machines. Thus, while
  Docker offers partial reproducibility: strong at run time (space) and weak at build time (time),
  it cannot on its own ensure verifiable, deterministic software artefacts. Achieving
  true reproducibility requires pinning image digests, freezing dependency versions,
  and employing deterministic build systems such as Nix or Guix for complete control
  over both space and time [@Malka_2024].

@@

I thank the rOpenSci reviewers and contributors who provided valuable feedback
on the development of [rix]{.pkg} and [rixpress]{.pkg}. In particular, I am
grateful to David Watkins and Jacob Wujiciak-Jens for their reviews of [rix]{.pkg},
and to William Landau and Anthony Martinez for their reviews of [rixpress]{.pkg}. I
also acknowledge the contributions of Philipp Baumann, [rix]{.pkg}’s co-author,
Richard J. Acton, Jordi Rosell, Elio Campitelli, László Kupcsik, and
Michael Heming for [rix]{.pkg}. Their expertise and feedback greatly
improved the quality and usability of these packages.

@@

This entire example can be executed by running the `run_polyglot_example.sh`
script available in the root of the paper's repository.
This script automates the full process:

1.  It first executes `gen-env.R` inside a temporary [Nix]{.pkg} shell to build the `default.nix` file.
2.  It then executes `gen-pipeline.R` inside the newly defined environment. This triggers [Nix]{.pkg} to run the
  entire polyglot pipeline in the correct order.

Upon completion, the script will have generated all intermediate artifacts and the final `readme.html` report found in
the `pipeline-output` folder containing the visualization.

>>>>

Other packages provide similar functionality with different trade-offs.
[groundhog]{.pkg} by @simonsohn2023 makes it easy to install packages as they
existed on CRAN at a given date:

```{=tex}
\begin{CodeInput}
R> groundhog.library("
+   library('purrr')
+   library('ggplot2')",
+   "2017-10-04",
+   tolerate.R.version = "4.2.2")
\end{CodeInput}
```

@@

5. Incomplete Reproducibility: Reproducibility should be understood as a
  spectrum rather than a binary property, a notion emphasised in
  "Reproducibility in Software Engineering" by @dellaieraMasterThesis2024.
  Docker illustrates this spectrum well: it provides strong run-time
  reproducibility—that is, consistency in space (where the same image behaves
  identically across different systems) but weak build-time reproducibility,
  which concerns consistency over time. As defined in the thesis,
  reproducibility at build time means being able to produce an identical build
  artefact regardless of when or where it is built. However, Docker inherently
  fails to guarantee this property. Its reliance on mutable base images and
  non-deterministic instructions such as `apt-get update` leads to temporal
  drift: rebuilding the same `Dockerfile` at different points in time can
  produce distinct outputs. In the experimental evaluation conducted in the
  thesis, even with pinned base images, checksums of the resulting containers
  varied across builds and machines. Thus, while Docker offers partial
  reproducibility: strong at run time (space) and weak at build time (time), it
  cannot on its own ensure verifiable, deterministic software artifacts.
  Achieving true reproducibility requires pinning image digests, freezing
  dependency versions, and employing deterministic build systems such as Nix or
  Guix for complete control over both space and time [@Malka_2024].

@@

I thank the rOpenSci reviewers and contributors who provided valuable feedback
on the development of [rix]{.pkg} and [rixpress]{.pkg}. In particular, I am
grateful to David Watkins and Jacob Wujiciak-Jens for their reviews of
[rix]{.pkg}, and to William Landau and Anthony Martinez for their reviews of
[rixpress]{.pkg}. I also acknowledge the contributions of Philipp Baumann,
[rix]{.pkg}’s co-author, Richard J. Acton, Jordi Rosell, Elio Campitelli, László
Kupcsik, and Michael Heming for [rix]{.pkg}. Their expertise and feedback
greatly improved the quality and usability of these packages. I would also like
to thank Pol Dellaiera for providing feedback on the manuscript.

@@

This entire example can be executed by running the `run_polyglot_example.sh`
script available in the root of the paper's repository.
This script automates the full process:
1.  It first executes `gen-env.R` inside a temporary [Nix]{.pkg} shell to build
    the `default.nix` file.
2.  It then executes `gen-pipeline.R` inside the newly defined environment. This
  triggers [Nix]{.pkg} to run the entire polyglot pipeline in the correct order.
Upon completion, the script will have generated all intermediate artifacts and
the final `readme.html` report found in the `pipeline-output` folder containing
the visualization.
